---
title: "xxx"
---

From Gavin Simpson's [gist](https://gist.github.com/gavinsimpson/bf11b821be44fc24e30ff8546ad8ca9d);
analyzing full transect data of J.A. Shultz et al (2016) PeerJ 4:e1980; DOI 10.7717/peerj.1980, [here](https://doi.org/10.7717/peerj.1980)


```{r}
library('readr')
library('dplyr')
library('tidyr')
## fitting stuff
library('lme4')
library('glmmTMB')
library('broom.mixed') 
## graphics stuff
library('ggplot2')
theme_set(theme_bw()) ## ggplot2 theme
library("ggstance")
```

Read in the data
```{r}
transects <- read_csv("transectcounts.csv")
(transects
        ## convert to long form
    %>% gather(species, count, -c(ssws,site,transect))
        ## convert to factors
    %>% mutate(ssws = factor(ssws, levels = c('before', 'after')),
               site = factor(site),
               transect = factor(transect))
) -> lt
nice_labels <- function(x) {
    ## https://stackoverflow.com/questions/15776732/how-to-convert-a-vector-of-strings-to-title-case
    re_from <- "\\b([[:lower:]])([[:lower:]]+)"
    x1 <- gsub(re_from, "\\U\\1\\L\\2", x, perl=TRUE)
    x2 <- gsub("\\."," ",x1)
    return(x2)
}
## desired species, in preferred order
spp_subset <- c("sunflower.star","green.urchin")
## subset down to the two species of interest, nicer/human-readable names
spp <- filter(lt, species %in% spp_subset) %>%
    mutate(species = factor(species,
                            levels = spp_subset,
                            labels = nice_labels(spp_subset)))
```

## explore

```{r dotplot}
gg1 <- ggplot(spp,aes(ssws,count))+
    geom_dotplot(binaxis="y",stackdir="center",dotsize=0.5,alpha=0.5)+
    facet_wrap(~species,scale="free",nrow=2)+
    stat_summary(fun.data=mean_cl_boot,colour="red",alpha=0.8)
```

```{r}
gg1 + scale_y_continuous(trans="log1p")+
```

```{r}
gg1 + scale_y_continuous(trans="sqrt")
```

But we need to account for transects. Each species/transect/ssws combination is measured once before and once after ...

```{r}
spp_tally <- spp %>% 
    spread(ssws,count) %>%
    group_by(species,before,after) %>%
    tally() %>%
    ungroup() %>%
    mutate(group=1:length(before)) %>%
    gather(ssws,count,-c(species,n,group)) %>%
    mutate(ssws=factor(ssws,levels=c("before","after")))
ggplot(spp_tally,aes(ssws,count))+
    geom_line(aes(size=n,group=group),alpha=0.1)+
    facet_wrap(~species,scale="free")+
    scale_y_continuous(trans="log1p")
```

This gets the pairwise relationship, but not the grouping of
transects within sites.

(As far as modeling goes, should be able to do a permutation test
(randomly swap before/after)? Do we need to account for grouping within
sites when computing summary statistics ... ?

## Model fitting - sunflower

```{r fit,cache=TRUE}
m.sun <- glmer(count ~ ssws + (ssws | site / transect),
               data = spp, family = poisson,
               subset = species == 'Sunflower Star')
m0.sun <- update(m.sun, . ~ . - ssws)
m.urchin <- update(m.sun, subset = species == 'Green Urchin')
m0.urchin <- update(m.urchin, . ~ . - ssws)
```

```{r tmp}
m.sun.pois <- glmmTMB(count ~ ssws + (ssws | site / transect),
               data = subset(spp, species == 'Sunflower Star'),
               family = poisson)
m.sun.pois.noobs <- update(m.sun.pois,
                           count ~ ssws + (1 | site / transect))
m.sun.nb <- update(m.sun.pois,
                   count ~ ssws + (1 | site / transect),
                   family = nbinom2)

m.sun.nb.ssws <- update(m.sun.nb,
                   count ~ ssws + (ssws | site) + (1 | site:transect))


m.sun.nb1 <- update(m.sun.nb, family = nbinom1)
m.sun.list <- lme4:::namedList(m.sun.pois,m.sun.nb,m.sun.nb1,
                               m.sun.nb.ssws,
                               m.sun.pois.noobs)
bbmle::AICtab(m.sun.list)
```

**Conclusions:** to my surprise, the Poisson (including an obs

Check zero-inflation:
```{r}
dat.sun <- subset(spp, species == 'Sunflower Star')
svals <- simulate(m.sun.pois,1000)
sumfun <- function(x) sum(x==0)
zdist <- apply(svals,2,sumfun)
plot(table(zdist))
zobs <- sumfun(dat.sun$count)
mean(zdist>zobs)
```

```{r}
fitted <- predict(m.sun,re.form=~1|site/transect,type="response")
resids <- (dat.sun$count-fitted)/sqrt(fitted)
plot(fitted,resids)
plot(fitted,dat.sun$count)
abline(a=0,b=1)
which(fitted<2 & dat.sun$count>10) ## obs 146
```

```{r diagnostics}
sum(residuals(m.sun,type="pearson")^2)/df.residual(m.sun)  ## underdispersed (?)
deviance(m.sun)/df.residual(m.sun)  ## similar result

library(broom.mixed)
aa <- augment(m.sun)
ggplot(aa,aes(.fitted,.resid))+geom_point() + geom_smooth()
ggplot(aa,aes(.fitted,sqrt(abs(.resid)),colour=site))+
          geom_point() + geom_smooth(aes(group=1))
## what's up with this? *seriously* heteroscedastic ...

spp_tmp <- (spp
    %>% filter(species=="Sunflower Star")
    %>% select(-species)
)

site_diff <- spp_tmp %>% spread(ssws,count) %>%
    group_by(site,transect) %>% summarise(diff=mean(before-after)) %>%
    arrange(diff)

spp_tmp <- spp_tmp %>%
    mutate(transect=factor(transect,levels=site_diff$transect))

ggplot(spp_tmp,aes(count,transect,colour=site,shape=ssws))+geom_point()

lattice::dotplot(ranef(m.sun,condVar=TRUE))$transect
```

Let's assume these are good models -
  should really do a lot of diagnostics...
```

## Extract data from the model for plotting

```{r}
ilink <- family(m.sun)$linkinv
effs.s <- tibble(species = 'Sunflower Star', ssws = c('before', 'after'), fitted = fixef(m.sun),
                 se = diag(vcov(m.sun)))
effs.g <- tibble(species = 'Green Urchin', ssws = c('before', 'after'), fitted = fixef(m.gur),
                 se = diag(vcov(m.gur)))

effs <- bind_rows(effs.s, effs.g)
effs <- mutate(effs, est = ilink(fitted), upper = ilink(fitted + (2*se)),
               lower = ilink(fitted - (2*se)),
               species = factor(species, levels = c('Sunflower Star','Green Urchin')),
               ssws = factor(ssws, levels = c('before', 'after')))

## base plot
base.plt <- ggplot() +
    facet_wrap(~ species, scales = 'free_y') +
    labs(x = 'Sea star mortality event')

## plot of effects
eff.plt <- base.plt +
    geom_pointrange(aes(y = est, x = ssws, ymax = upper, ymin = lower), data = effs) +
    labs(y = 'Expected count')
eff.plt

ggsave('model-effects.png', eff.plt, width = 7, height = 3.5)

## plot of effects on data
data.plt <- base.plt +
    geom_point(aes(y = count, x = ssws), data = spp, alpha = 0.2,
               position = position_jitter(width = 0.2)) +
    geom_pointrange(aes(y = est, x = ssws, ymax = upper, ymin = lower, colour = ssws),
                    data = effs) +
    theme(legend.position = 'none') +
    labs(y = 'Number of individuals')
data.plt

ggsave('data-plus-effects-original-scale.png', data.plt, width = 7, height = 3.5)

## Data on sqrt scale
data.plt + scale_y_sqrt(minor_breaks = NULL) +
    expand_limits(y = 0)

ggsave('data-plus-effects-sqrt-scale.png', width = 7, height = 3.5)
```
