---
title: "Interactive Graphics"
author: "Angela Wang & Jing Cai"
date: "3/1/2018"
output:
 slidy_presentation:
 incremental: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
```

# <br> Some amazing dynamic graphs...
<br>
<br>
1,000 gold medals in the Summer Olympics of team USA:  
 <https://www.washingtonpost.com/graphics/sports/olympics/the-1000-medals-of-the-united-states/>  
Gapminder world poster:  
 <https://www.gapminder.org/tools/#_chart-type=bubbles>  
Visual proof of why vaccines do more good than harm:  
 <http://www.sciencemag.org/news/2017/04/here-s-visual-proof-why-vaccines-do-more-good-harm>  

# <br> Why interactive graphs?  
<br>
<br>
* Allows users to search for information quickly without fully specified questions  
* Engages the users, thus aids communication and enhances knowledge transfer  
* "Any high-dimensional dataviz has to be summarized in some way, but interactivity allows us to get details" --- Dr. Karl Broman  
  
  
  
# <br> Different types of interactions  
<br>
<br>
* Plot specific interaction  
* Scaling: zoom in and out  
* Identification: display a label with detailed information  
* Brushing: highlight or hide selected variables  
* Linking: connect elements selected in all plots  
  
  
  
# <br> Some interactive visualization packages in R
<br>
<br>

Package     | Description
----------- | -------------------------------------------------------------
ggvis       | interactive plots from the maker of ggplot2  
plotly      | convert ggplot2 figures to interactive plots easily   
googleVis   | use Google Chart Tools from R  
highcharter | R interface for highcharts, a javascript charting library  
crosstalk   | extends htmlwidgets with a set of classes, functions, and conventions for implementing cross-widget interactions   

# Example - dataset
```{r}
dat1<-(read_csv("stock_price.csv")
       %>% select(c(Company,Date,Close,Volume)))
head(dat1)
```

# Example - plot using ggplotly
```{r}
gg1<-(ggplot(dat1,aes(x=Volume,y=Close,colour=Company))+
        geom_line()
)
ggp1<-ggplotly(gg1)
ggp1
```
  
  
  
***
<br>
<br>
<br>
<br>
<br> 
<br> 
<center><h3> The greatest value of a picture is when it forces us to notice what we never expected to see.</h3><center>  
<center><h3> ---John Tukey</h3><center>  
