---
title: "tidyverse/ggplot example"
---

```{r pkgs,message=FALSE}
library(ggplot2)
## I have a strong preference for the "black & white" theme
##   (white background, gray gridlines)
theme_set(theme_bw())
library(readr)   ## tidyverse file-I/O
library(dplyr)   ## filter/select/mutate/etc.
library(tidyr)   ## reshaping
library(GGally)  ## pairs plots
```

```{r get_cdat}
cdat <- read_csv("../data/Carbohydrate_diet.csv")
## tibble does magic: <= 20 rows, prints all by default;
##    > 20 rows, prints first 10 by default
print(cdat,n=8)
```

Univariate plot:
```{r univ1}
gg0 <- (ggplot(cdat,aes(weight,carbohydrate))
    + geom_point()
)
```

Add two different types of smooth line:
```{r univ_multismooth}
print(gg0
      + geom_smooth() ## default:loess
      + geom_smooth(method="lm",col="red",fill="red") ## linear model
      )
```

This data set has no categorical variable, so we'll make an
age-category variable:
```{r mkcat}
cdat <- mutate(cdat,f_age=cut_number(age,3))
```

```{r grp_regression}
gg1 <- (ggplot(cdat,aes(weight,carbohydrate,colour=f_age))
    + geom_point(aes(shape=f_age),size=4)
)
gg2 <- (gg1
    + geom_smooth(method="lm",
                  aes(color=f_age,fill=f_age))
)
print(gg2)
```

Categorical variable on x axis:

```{r boxplot}
ggplot(cdat,aes(f_age,carbohydrate)) + geom_boxplot()+
    geom_point(size=4,alpha=0.2)
```
(also see `geom_jitter()`)

Faceting:
```{r facet1}
gg3 <- (gg2
    + facet_wrap(~f_age,labeller=label_both)
)
print(gg3)
```

Squash panels together:
```{r facet2}
z_margin <- theme(panel.spacing=grid::unit(0,"lines"))
print(gg3 + z_margin)
```

## Univariate analyses

Collapse data by variable:
```{r gather}
(cdat %>%
 select(-f_age) %>%
 gather(key=pvar,value=value,
        -carbohydrate ## don't include carbohydrate
        ) 
) -> cdat_m
```

Plot univariate regressions:

```{r univ_reg,fig.width=8}
ggplot(cdat_m,aes(x=value,y=carbohydrate))+
    geom_point()+facet_wrap(~pvar,scale="free")+
    geom_smooth()+
    geom_smooth(method="lm",colour="red",fill="red")
```

Finally, a fancy pairs plot:

```{r pairs,message=FALSE}
ggpairs(cdat)
```

## contraception data

Binary response data: challenging!

```{r contr1}
data("Contraception",package="mlmRev")
gg0 <- ggplot(Contraception,aes(age,use))+geom_point()
print(gg0)
```

Collapse data:
```{r contr_sum}
(Contraception
    ## collapse age by year
    %>% mutate(f_age=cut(age,breaks=(-14):20,labels=seq(-13.5,19.5)),
         ## then turn it back into a number
               f_age=as.numeric(as.character(f_age)))
    ## means by age group/urban vs rural/number of live children
    %>% group_by(f_age,urban,livch)
    ## compute proportion, n, standard error
    %>% summarise(prop=mean(as.numeric(use)-1),
                  n=n(),
                  se=sqrt(prop*(1-prop)/n))
) -> contr_sum
```

```{r contr_sum_plot}
gg1 <- ggplot(contr_sum,aes(f_age,prop,colour=urban)) +
    geom_point(aes(size=n))+
    geom_linerange(aes(ymin=prop-1*se,ymax=prop+1*se))+
    facet_wrap(~livch,labeller=label_both)+
    scale_colour_brewer(palette="Dark2")
print(gg1)
```

add quadratic logistic regression curves ...

```{r contr_sum_reg_plot}
gg1 + geom_smooth(data=Contraception,
                  inherit.aes=FALSE, ## not sure why we need this?
                  aes(x=age,y=as.numeric(use)-1,colour=urban),
                  method="glm",
                  method.args=list(family=binomial),
                  formula=y~poly(x,2))
```

```{r}
ggplot(Contraception,aes(age,as.numeric(use)-1))+
    geom_point()+geom_smooth(method="glm",aes(group=1),
                             method.args=list(family=binomial),
                             formula=y~poly(x,2))

ggplot(Contraception,aes(age,as.numeric(use)-1))+
    geom_point()+geom_smooth(method="gam",aes(group=1),
                             method.args=list(family=binomial),
                             formula=y~s(x,k=40))
```
