---
title: "NIMBLE + Wrapup"
author: "Mike Li, Morgan Kain"
date:  "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output: html_document
---

```{r setup}
library(knitr)
opts_knit$set(root.dir = "/Users/Morgan/Documents/Research/McMaster/Modeling_Work/Mike_Lunchbox/lunchbox")
#opts_chunk$set(cache = TRUE)
```

Our goals for the final presentation are the following: <br />
1. Describe the use of a powerful tool for MCMC and other applications (Particle Filtering, EM) <br />
2. Present the steps of a validation process using a variety of methods/platforms <br />
3. We do not have much new "Thursday Material" so the majority of our time will be spent in R <br />

## NIMBLE: Numerical Inference for statistical Models for Bayesian and Likelihood Estimation

NIMBLE is built in R but compiles your models and algorithms using C++ for speed <br />
NIMBLE is most commonly used for MCMC but can also be used to implement particle filtering and EM
1. A system for writing statistical models flexibly, which is an extension of the BUGS language <br />
2. A library of algorithms such as MCMC. <br />
3. A language, called NIMBLE, embedded within and similar in style to R, for writing <br />
  algorithms that operate on BUGS models.
  <br />
  
One of the most important concepts behind NIMBLE is to allow a combination of highlevel 
processing in R and low-level processing in compiled C++. <br />

1. NIMBLE extracts all the declarations in the BUGS code to create a model definition. <br />
2. From the model definition, NIMBLE builds a working model in R. This can be used
to manipulate variables and operate the model from R. Operating the model includes
calculating, simulating, or querying the log probability value of model nodes. <br />
3. From the working model, NIMBLE generates customized C++ code representing the
model, compiles the C++, loads it back into R, and provides an R object that interfaces
to it. We often call the uncompiled model the “R-model” and the compiled model the
“C-model.” <br />  

##### Why NIMBLE?

1. Options (More customizable MCMC, ability to run JAGS models and STAN models, EM, particle filter) <br />
2. User-defined functions and distributions – written as nimbleFunctions – can be used
in model code.  <br />
3. Multiple parameterizations for distributions, similar to those in R, can be used. <br />
  e.g. normal distribution with BUGS parameter order: <br />
        x ~ dnorm(a + b * c, tau) <br />
       normal distribution with a named parameter: <br />
        y ~ dnorm(a + b * c, sd = sigma) <br />
4. Named parameters for distributions and functions, similar to R function calls, can be
used. <br />
5. More flexible indexing of vector nodes within larger variables is allowed. For example
one can place a multivariate normal vector arbitrarily within a higher-dimensional
object, not just in the last index. <br />
6. More general constraints can be declared using dconstraint, which extends the concept
of JAGS’ dinterval. <br />


#### Downloading, installing and loading NIMBLE

On Windows, you should download and install Rtools.exe available from http://cran. r-project.org/bin/windows/Rtools/.
On OS X, you should install Xcode.
You wil also need the make utility

After these are installed you can install NIMBLE in R using <br />
install.packages("nimble", repos = "http://r-nimble.org", type = "source") <br />

Please post about installation problems to the nimble-users Google group or email nimble.stats@gmail.com.

You will also need to download STAN using the following commands <br />
Sys.setenv(MAKEFLAGS = "-j4") <br />
install.packages("rstan", dependencies = TRUE) <br />

In total you will need the following pakages:

```{r loadpackage, echo=TRUE, message=FALSE, verbose = FALSE}
library("nimble")
library("R2jags")
library("ggplot2")
library("nimble")
library("rstan")
library("igraph")
library("parallel")
library("mcmcplots")
library("lattice")
```

#### NIMBLE Overview

<b> Programming in NIMBLE involves a fundamental distinction between: </b> <br />
  1. the steps for an algorithm that need to happen only once, at the beginning, such as inspecting the model <br />
  2. the steps that need to happen each time a function is called, such as MCMC iterations. <br />
    When one writes a nimbleFunction, each of these parts can be provided separately. 

Multiple parameterizations for distributions, similar to those in R, can be used.
NIMBLE calls non-stochastic nodes “deterministic”, whereas BUGS calls them “logical”. 
NIMBLE uses “logical” in the way R does, to refer to boolean (TRUE/FALSE) variables.
Alternative models can be defined from the same model code by using if-then-else statements that are evaluated when the model is defined.

### Presentation Outline
The general outline for this presentation follows along with the NIMBLE users manual <br />
http://r-nimble.org/documentation-2 <br />
However, the model(s) used here are written by us <br />

##### Part 1
[1.1](#1.1) Build a chain binomial model in JAGS. Conduct parameter estimation <br />
[1.2](#1.2) Translate the model into NIBLE. Conduct parameter estimation <br />
      [1.2.1](#1.2.1) Model exploration/conversion <br />
      [1.2.2](#1.2.2) Create a basic MCMC specification for the chain binomial, compile and run the MCMC <br />
      [1.2.3](#1.2.3) Small MCMC specification adjustment (more on this in Part 2) <br />
[1.3](#1.3) Compare the JAGS and NIMBLE results (parameter estimates, uncertainty, convergence, computation time) <br />
[1.4](#1.4) Translate the model using a "hybrid approach" (STAN does not allow for discrete latent variables) <br />
      [1.4.1](#1.4.1) Conduct parameter estimation using JAGS and NIMBLE <br />
      [1.4.2](#1.4.2) Run the hybrid model in STAN and compare the results from JAGS, NIMBLE and STAN <br />
[1.5](#1.5) Compare the NIMBLE Chain Binomial and STAN hybrid model <br />

##### Part 2
[2.1](#2.1) Expolore more fine-tuned adjustments that can be made in NIMBLE <br />
      [2.1.1](#2.1.1)  MCMC specifications <br />
      [2.1.2](#2.1.2)  NIMBLE functions (e.g. allows for the implementation of custom samplers) <br />
      
##### Part 3     
[3.1](#3.1) NIMBLE extras: <br />
      [3.1.1](#3.1.1) Create, compile and run a Monte Carlo Expectation Maximization (MCEM) algorithm, which illustrates some of the flexibility NIMBLE provides to combine R and NIMBLE. <br />
      [3.1.2](#3.1.2) Implement particle filtering for the chain binomial <br />
      
##### Part 4

[4.1](#4.1) Misc NIMBLE notes (truncated distributions, lifted nodes, logProb)

### Part 1

##### <a name="1.1"> 1.1 Build a chain binomial model in JAGS </a>
First step is to construct the simulator from which we will obtain our data <br />

Note: It will be important to set your current working directory to "../stat744/notes/NIMBLE" <br />

Set parameters and load the Chain Binomial simulator <br />

```{r, cache.rebuild = TRUE}
beta <- 0.02
pop <- 100
effpropS <- 0.8
effpropI <- 0.2
reporting <- 0.5

s0 <- effpropS*pop
r0 <- 0
zerohack <- 0.001
numobs <- 12
nimtimevec <- c()
source("CBsimulator.R")
```

```{r, cache.rebuild = TRUE}
sim <- simCB(beta = beta, pop = pop, effpropS = effpropS, effpropI = effpropI, 
             t0 = 1, numobs = numobs, reporting = reporting, seed = 3)
sim
```

Take a peek at what this model produces

```{r, echo=FALSE}
  ggplot(data.frame(sim), aes(time, S)) + geom_point(colour = "blue") + 
    geom_line(colour = "blue") +
    geom_line(data = data.frame(sim), aes(time, I), colour = "red") +
    geom_line(data = data.frame(sim), aes(time, R), colour = "green") +
    theme_bw()
```

Set up the required arguments to run the JAGS model

```{r, tidy = TRUE}
data <- list(obs = sim$Iobs,
             pop = pop,
             numobs = nrow(sim),
             r0 = r0)

inits <- list(list(
  I = sim$I*1 + 1,
  effpropS = effpropS,
  effpropI = effpropI,
  beta = beta,
  reporting = reporting),
list(
  I = sim$I*1 + 1,
  effpropS = effpropS,
  effpropI = effpropI,
  beta = beta,
  reporting = reporting),
list(
  I = sim$I*1 + 1,
  effpropS = effpropS,
  effpropI = effpropI,
  beta = beta,
  reporting = reporting)
)

params = c("beta",
           "effpropS",
           "effpropI",
           "reporting")

#rjags::set.factory("bugs::Conjugate", FALSE, type="sampler")
```

Create the model and examine the MCMC algorithms that JAGS will use to sample <br />

```{r}
cbjagsmodel <- jags.model(data = data,
               inits = inits,
               file = "CB.bug",
               n.chains = length(inits))

list.samplers(cbjagsmodel)
```

Run some chains (could use coda::coda.samples from cbjagsmodel but here we will just run jags()) <br />

```{r}
jagstime <- system.time(cbjags <- jags(data = data,
               inits = inits,
               param = params,
               model.file = "CB.bug",
               n.iter = 11000,
               n.burnin = 500,
               n.thin = 20,
               n.chains = length(inits)))
```

```{r}
cbjags
xyplot(as.mcmc(cbjags))
```

##### <a name="1.2"> 1.2 Build the NIMLE model </a>

```{r}
source('nimCB.R')
```

Set up the model. Here we need: Constants, Data, Initial Values, NIMBLE model object <br />

```{r}
nimCBdata <- list(obs = sim$Iobs)

nimCBcon <- list(numobs = numobs, pop = pop, r0 = r0)

nimCBinits <- list(I = sim$I,
                   effpropS = effpropS,
                   effpropI = effpropI,
                   beta = beta,
                   reporting = reporting,
                   s0 = s0)

nimtimevec[1] <- system.time(CBout <- nimbleModel(code = nimcode, 
                         name = 'CBout', 
                         constants = nimCBcon,
                         data = nimCBdata, 
                         inits = nimCBinits))[3]
```

##### <a name="1.2.1"> 1.2.1 Model exploration/conversion </a>

```{r}
CBout$getNodeNames()
```

```{r}
CBout$obs
```

```{r}
par(mfrow = c(1,1))
plot(CBout$graph)
```

nimbleModel does its best to initialize a model, but let’s say you want to re-initialize I. <br />

```{r}
simulate(CBout, 'I')
CBout$I
```

```{r}
CBout$getDependencies(c("beta", "effpropS", "effpropI", "reporting"))
```

```{r}
CBout$getDependencies(c("beta", "effpropS", "effpropI", "reporting"), determOnly = TRUE)
```

##### <a name="1.2.2"> 1.2.2 Compile the C++ model, build the MCMC and Run </a>

```{r}
nimtimevec[2] <- system.time(CBoutC <- compileNimble(CBout))[3]
```

Configure the MCMC with the default options (we will return to customizing this setup later) <br />

```{r, message=FALSE, verbose = FALSE, warning = FALSE, errors = FALSE}
nimtimevec[3] <- system.time(CBoutSpec <- configureMCMC(CBout, print = TRUE))[3]
```

Add chain monitors for the parameters of interest <br />

```{r}
CBoutSpec$addMonitors(c("beta", "effpropS", "effpropI", "reporting"))
```

Build the MCMC <br />

```{r}
nimtimevec[4] <- system.time(CBoutMCMC <- buildMCMC(CBoutSpec))[3]
nimtimevec[5] <- system.time(CBoutMCMC <- compileNimble(CBoutMCMC, project = CBout, resetFunctions = TRUE))[3]
```

```{r}
niter <- 11000
set.seed(0)
nimtimevec[6] <- system.time(CBoutMCMC$run(niter))[3]
```

Quick peek at time required

```{r}
jagstime[3]
sum(nimtimevec[1:6], na.rm = TRUE)
nimtimevec[6]
```

```{r, echo = FALSE}
  samples <- as.matrix(CBoutMCMC$mvSamples)
  par(mfrow = c(1, 2), mai = c(.6, .5, .1, .2))
  plot(samples[ , 'beta'], type = 'l', xlab = 'iteration')
  plot(samples[ , 'effpropS'], type = 'l', xlab = 'iteration')
  plot(samples[ , 'effpropI'], type = 'l', xlab = 'iteration')
  plot(samples[ , 'reporting'], type = 'l', xlab = 'iteration')
  plot(samples[ , 'effpropS'], samples[ , 'effpropI'])
```

Look at the correlation in the chains

```{r}
  acf(samples[, "beta"])
  acf(samples[, "reporting"])
  acf(samples[, "effpropS"])
  acf(samples[, "effpropI"])
```

##### <a name="1.2.3"> 1.2.3 Small MCMC specification adjustment </a>

A few undesirable results here... we can add a block sampler to decrease correlation <br />

```{r}
CBoutSpec$addSampler(target = c('effpropS', 'effpropI'), type = 'RW_block',
                      control = list(adaptInterval = 10000))
```

```{r}
CBoutSpec$setThin(25)
```

```{r}
CBoutMCMC <- buildMCMC(CBoutSpec)
```

```{r}
CBoutMCMC <- compileNimble(CBoutMCMC, project  = CBout, resetFunctions = TRUE)
```

```{r}
CBoutMCMC$run(20000)
samplesNew <- as.matrix(CBoutMCMC$mvSamples)
```

Check for an imporvement

```{r}
  par(mfrow = c(2,2))
  acf(samplesNew[, "effpropS"])
  acf(samplesNew[, "effpropI"])
  plot(samplesNew[ , 'effpropS'], type = 'l', xlab = 'iteration')
  plot(samplesNew[ , 'effpropI'], type = 'l', xlab = 'iteration')
  par(mfrow = c(1,1))
  plot(samplesNew[ , 'effpropS'], samplesNew[ , 'effpropI'])
```

##### <a name="1.3"> 1.3 Compare the JAGS and NIMBLE results </a>

We can also compare the NIMBLE model simultaneously with the JAGS model using MCMCsuite() <br />

Be warned: running this code will produce about 6-8 graphs which will all pop up in separate windows! <br />

```{r}
nimcb <- MCMCsuite(code = nimcode,
                   data = nimCBdata,
                   inits = nimCBinits,
                   constants = nimCBcon,
                   MCMCs = c("jags", "nimble"),
                   monitors = c("beta", "reporting", "effpropS", "effpropI"),
                   niter = 12000,
                   makePlot = FALSE,
                   savePlot = FALSE)
```

![beta chains](/Users/Morgan/Documents/Research/McMaster/Modeling_Work/Mike_Lunchbox/lunchbox/beta_1.pdf)
![report chains](/Users/Morgan/Documents/Research/McMaster/Modeling_Work/Mike_Lunchbox/lunchbox/reporting_1.pdf)
![posterior distributions](/Users/Morgan/Documents/Research/McMaster/Modeling_Work/Mike_Lunchbox/lunchbox/density_1.pdf)

##### <a name="1.4"> 1.4 "hybrid approach" </a> 

We must rewrite the model so that there are no discrete latent variables. We call this the "hybrid model" <br />
An asside -- Discrete Latent Variables: <br />
An additional asside -- Hamiltonian MCMC: <br />

But before we fit the model in STAN lets explore the hybrid model in NIMBLE <br />

NIMBLE allows us to compare the results of multiple models even if they have different parameterizations 
(e.g. Chain Binomial and the Hybrid Model) <br />

```{r}
data$obs <- data$obs + zerohack # Guarnantee that obs remains above 0 (important for the gamma)
data$zerohack <- zerohack

hybridjags <- jags(data = data,
               inits = inits,
               param = params,
               model.file = "hybrid.bug",
               n.iter = 8000,
               n.chains = length(inits))
```

##### <a name="1.4.1"> 1.4.1 Hybrid in JAGS and NIMBLE </a>

```{r}
source('nimhybrid.R')
```

```{r}
nimhydata <- list(obs = sim$Iobs + zerohack)
nimhycon <- list(numobs = numobs, pop = pop, r0 = r0, zerohack = zerohack)

nimhyinits <- list(I = sim$I + zerohack,
                   effpropS = effpropS,
                   effpropI = effpropI,
                   beta = beta,
                   reporting = reporting,
                   s0 = s0)
```

```{r}
nimcb <- MCMCsuite(code = nimcode,
                   data = nimhydata,
                   inits = nimhyinits,
                   constants = nimhycon,
                   MCMCs = c("jags", "nimble"),
                   monitors = c("beta", "reporting", "effpropS", "effpropI"),
                   niter = 10000,
                   makePlot = FALSE,
                   savePlot = FALSE)
```

##### <a name="1.4.2"> 1.4.2 Hybrid in JAGS, NIMBLE and STAN </a>

Run the STAN model <br />

```{r, message=FALSE, verbose = FALSE, warning = FALSE, errors = FALSE}
stantime <- system.time (s1 <- stan(file='hybrid.stan', data = data, init = inits,
           pars=c("beta", "reporting", "effpropS", "effpropI", "I"), iter = 8000,
           seed = 1001, chains = length(inits)))
```

Compare all three methods using the hybrid model <br />

```{r}
nimhydata <- list(obs = sim$Iobs + zerohack)
nimhycon <- list(numobs = numobs, pop = pop, 
                 r0 = r0, zerohack = zerohack)

nimhyinits <- list(I = sim$I + zerohack,
                   effpropS = effpropS,
                   effpropI = effpropI,
                   beta = beta,
                   reporting = reporting,
                   s0 = s0)
```

```{r}
allhybrids <- MCMCsuite(code = nimcode,
                   data = nimhydata,
                   inits = nimhyinits,
                   constants = nimhycon,
                   stan_model = "hybrid.stan",
                   MCMCs = c("jags", "nimble", "stan"),
                   monitors = c("beta", "reporting", "effpropS", "effpropI"),
                   niter = 10000,
                   makePlot = FALSE,
                   savePlot = FALSE)
```

##### <a name="1.5"> 1.5 Finally, compare the Chain Binomial NIMBLE and Hybrid STAN </a>

```{r}
nimCBdata <- list(obs = sim$Iobs)
nimCBcon <- list(numobs = numobs, pop = pop, r0 = r0, zerohack = zerohack)

nimCBinits <- list(I = sim$I,
                   effpropS = effpropS,
                   effpropI = effpropI,
                   beta = beta,
                   reporting = reporting,
                   s0 = s0)
```

```{r, eval = FALSE}
nimcb <- MCMCsuite(code = nimcode,
                   data = nimCBdata,
                   inits = nimCBinits,
                   constants = nimCBcon,
                   stan_model = "hybrid.stan",
                   MCMCs = c("jags", "nimble", "stan"),
                   monitors = c("beta", "reporting", "effpropS", "effpropI"),
                   niter = 10000,
                   makePlot = TRUE,
                   savePlot = TRUE)
```

### Part 2

See pg 53-55 for how nodes are named <br />

See pg 70-85 for samplers <br />

See pg 87 for custom MCMC sampler <br />

```{r}

```

```{r}

```

```{r}

```

```{r}

```
```{r}

```

```{r}

```
```{r}

```

```{r}

```

### Part 3

##### Mote Carlo Expectation Maximization

Suppose we have a model with missing data (or a layer of latent variables that can be 
treated as missing data) and we would like to maximize the marginal likelihood of the model,
integrating over the missing data. A brute-force method for doing this is MCEM. <br />

Start by building the model <br />

```{r, eval = FALSE}
hybemout <- nimbleModel(code = nimcode, 
                         name = 'hybemout', 
                         constants = nimhycon,
                         data = nimhydata, 
                         inits = nimhyinits)

hybMCEM <- buildMCEM(model = hybemout, latentNodes = list("I", "s0"), 
                      burnIn = 100, mcmcControl = list(adaptInterval = 20), 
                      boxConstraints = list( list( c("beta", "reporting", "effpropS", "effpropI"), 
                                                   limits = c(0, Inf) ) ), 
                        buffer = 1e-6)

# The MCEM algorithm allows for box constraints on the nodes that will be optimized,
# specified via the boxConstraints argument. This is highly recommended for nodes that
# have zero density on parts of the real line
```

```{r, eval = FALSE}
CBMCEM(maxit = 20, m1 = 150, m2 = 200)
```

##### Particle Filter 

Set up the Nimble model <br />

```{r}
CBpfout <- nimbleModel(code = nimcode, 
                         name = 'CBpfout', 
                         constants = nimCBcon,
                         data = nimCBdata,
                         check = FALSE)
```

Build the particle filter <br />

```{r}
CBpfoutC <- compileNimble(CBpfout)

CBpf <- buildPF(CBpfout, c("I"))
```

```{r}
CBpfC <- compileNimble(CBpf, project = CBpfout)
```

Set your parameters <br />

```{r}
CBpfoutC$beta = 0.02
CBpfoutC$effpropS = 0.8
CBpfoutC$effpropI = 0.2
CBpfoutC$reporting = 0.5
```

Obtain log-likelihood

```{r}
CBpfC$run(m = 5000)
```

Currently relatively useless as is... <br />

Use this framework to construct your own updater <br />

##### Part 4:  

<a name="4.1"> NIMBLE Notes </a>

**Truncation of distributions** <br />
   • $x \∼ N(0, sd = 10) T(0, a)$, or <br />
   • x ∼ T(dnorm(0, sd = 10), 0, a), <br />

   • mu1 ~ dnorm(0, 1) <br />
   • mu2 ~ dnorm(0, 1) <br />
   • constraint_data ~ dconstraint( mu1 + mu2 > 0 ) <br />
   
**Lifted Nodes** <br />
   • The use of link functions causes new nodes to be introduced <br />
   • When distribution parameters are expressions, NIMBLE creates a new deterministic node that contains the expression for a given parameter <br />
   
**logProb** <br />
    • For each variable that contains at least one stochastic node, NIMBLE generates a model variable with the prefix “logProb” <br />
    • Can be retrieved with getLogProb
    
**Choice of Samplers** <br /> 
    1. If the node has no stochastic dependents, a predictive end sampler is assigned. The end sampling algorithm merely calls simulate on the particular node. <br /> 
    2. The node is checked for presence of a conjugate relationship between its prior distribution and the distributions of its stochastic dependents. If it is determined to be in a conjugate relationship, then the corresponding conjugate (Gibbs) sampler is assigned. <br /> 
    3. If the node is discrete-valued, then a slice sampler is assigned [5]. <br /> 
    4. If the node follows a multivariate distribution, then a RW block sampler is assigned for all elements. This is a Metropolis-Hastings adaptive random-walk sampler with a multivariate normal proposal [6]. <br /> 
    5. If none of the above criteria are satisfied, then a RW sampler is assigned. This is a Metropolis-Hastings adaptive random-walk sampler with a univariate normal proposal distribution. <br /> 
    
**Missing Values** <br /> 
    See pg 43