---
title: "NIMBLE + Wrapup"
author: "Mike Li, Morgan Kain"
date:  "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output: html_document
---

Our goals for the final presentation are the following: <br />
1. Describe the use of a powerful tool for MCMC and other applications (Particle Filtering, EM) <br />
2. Present the steps of a validation process using a variety of methods/platforms <br />
3. We do not have much new "Thursday Material" so the majority of our time will be spent in R <br />

## NIMBLE: Numerical Inference for statistical Models for Bayesian and Likelihood Estimation

NIMBLE is built in R but compiles your models and algorithms using C++ for speed <br />
NIMBLE is most commonly used for MCMC but can also be used to implement particle filtering and EM
1. A system for writing statistical models flexibly, which is an extension of the BUGS language <br />
2. A library of algorithms such as MCMC. <br />
3. A language, called NIMBLE, embedded within and similar in style to R, for writing <br />
  algorithms that operate on BUGS models.
  <br />

#### Downloading, installing and loading NIMBLE

On Windows, you should download and install Rtools.exe available from http://cran. r-project.org/bin/windows/Rtools/.
On OS X, you should install Xcode.
You wil also need the make utility

Please post about installation problems to the nimble-users Google group or email nimble.stats@gmail.com.

After these are installed you can install NIMBLE in R using <br />
install.packages("nimble", repos = "http://r-nimble.org", type = "source") <br />

You will also need to download STAN using the following commands <br />
Sys.setenv(MAKEFLAGS = "-j4") <br />
install.packages("rstan", dependencies = TRUE) <br />

In total you will need the following pakages:

```{r loadpackage, echo=TRUE, message=FALSE}
library("nimble")
library("R2jags")
library("ggplot2")
library("nimble")
library("rstan")
```

#### NIMBLE Overview

<b> Programming in NIMBLE involves a fundamental distinction between: </b> <br />
  1. the steps for an algorithm that need to happen only once, at the beginning, such as inspecting the model <br />
  2. the steps that need to happen each time a function is called, such as MCMC iterations. <br />
    When one writes a nimbleFunction, each of these parts can be provided separately. 

Multiple parameterizations for distributions, similar to those in R, can be used.
NIMBLE calls non-stochastic nodes “deterministic”, whereas BUGS calls them “logical”. 
NIMBLE uses “logical” in the way R does, to refer to boolean (TRUE/FALSE) variables.
Alternative models can be defined from the same model code by using if-then-else statements that are evaluated when the model is defined.

#### Presentation Outline
The general outline for this presentation follows along with the NIMBLE users manual <br />
http://r-nimble.org/documentation-2 <br />
However, the model used here is a simple SIR model written by us <br />

1. Build an SIR model in JAGS. Conduct parameter estimation and forecasting <br />
2. Translate the model into NIBLE. Conduct parameter estimation and forecasting <br />
      2.1 Model conversion <br />
      2.2 Compile the model <br />
      2.3 Create a basic MCMC specification for the pump model <br />
      2.4 Compile and run the MCMC <br />
3. Compare the results between these two approaches (parameter estimates, uncertainty, convergence, computation time) <br />
      3.1 Customize the MCMC specification and compile and run that <br />
4. Write the same model in STAN. Conduct parameter estimation and forecasting <br />
5. Translate this new model into NIMBLE. <br />
6. Compare all results <br />
7. Create, compile and run a Monte Carlo Expectation Maximization (MCEM) algorithm, which illustrates some of the flexibility NIMBLE provides to combine R and NIMBLE. <br />
8. Implement particle filtering for this same model <br />
9. Write a short nimbleFunction to generate simulations from designated nodes of any model. <br />

##### Build an SIR model in JAGS
First step is to construct the simulator from which we will obtain our data

```{r}
simi <- function(beta, gamma, pop, i0,
                 t0, end, dt, report, seed){
  if (!is.null(seed)) set.seed(seed)
  tvec <- seq(1, end, by = dt)
  n <- length(tvec)
  I <- S <- pI <- infobs <- foi <- rec <- inf <- numeric(end)
  
  ## Initial conditions
  I[1] <- rnbinom(1, 1, 1/4) + 1
  S[1] <- pop - I[1]
  pR <- 1 - exp(-gamma)
  pI[1] <- 1 - exp(-beta*I[1]*S[1]/pop)
  inf[1] <- rbinom(1, prob = pI[1], size = S[1])
  infobs[1] <- rbinom(1, prob = report, size = inf[1])
  rec[1] <- rbinom(1, prob = pR, size = I[1])
  
  ## Generate the Unobserved process I, and observables
  
  for (t in 2:end){
    I[t] <- I[t - 1] + inf[t - 1] - rec[t - 1]
    S[t] <- S[t - 1] - inf[t - 1]
    pI[t] <- 1 - exp(-beta*I[t]*S[t]/pop)
    inf[t] <- rbinom(1, prob = pI[t], size = S[t])
    infobs[t] <- rbinom(1, prob = report, size = inf[t])
    rec[t] <- rbinom(1, prob = pR, size = I[t])
  }
  
  cbind(tvec, S, I, inf, infobs)
}
```

```{r, echo=FALSE}
beta <- 0.00015; pop <- 5000; gamma = 0.04; end = 100; report = 0.8; nchain = 3

SIRdata <- simi(beta = beta, pop = pop, gamma = gamma,
                       end = end, report = report, seed = 482)
```

Take a peek at what this model produces

```{r, echo=FALSE}
  ggplot(data.frame(SIRdata), aes(tvec, S)) + geom_point(colour = "blue") + 
    geom_line(colour = "blue") +
    geom_point(data = data.frame(SIRdata), aes(tvec, I), colour = "red") +
    geom_line(data = data.frame(SIRdata), aes(tvec, I), colour = "red") +
    theme_bw()
```



```{r}
  jagsdat <- list(infobs = SIRdata[, 5], pop = pop, end = end, 
                  i0 = SIRdata[1, 3])

  parameters <- c("beta", "gamma", "report")
  
inits <- list(list(
	beta = beta/10, 
	report = 0.2,
	inf = SIRdata[, 5]*1 + 1
), list(
	beta = beta, 
	report = 0.3,
	inf = SIRdata[, 5]*1 + 1
), list(
	beta = beta*10, 
	report = 0.4,
	inf = SIRdata[, 5]*1 + 1
))
```

Place 'jagsSIRmodel.bug' in your working directory <br />

##### Note: Many of these builds take a very long time to run. Feel free to adjust "n.iter" or "end" or load the .rda file included

Create the model and examine the MCMC algorithms that JAGS will use to sample <br />

```{r}
 SIRmodel <- jags.model(file = "jagsSIRmodel.bug", data = jagsdat, inits = inits, 
             n.chains = length(inits), n.adapt = 1000)

list.samplers(SIRmodel)
```

Run some chains
```{r}
SIRsamples <- jags(data=jagsdat, inits=inits, param=parameters, 
             model.file = SIR.jagsmodel,   
             n.chains=length(inits), n.iter=5000, n.burnin=500, 
             n.thin=20, 
             refresh = 100)
```

```{r}
SIRsamples
mcmcplot(SIRsamples)
```

Very poor mixing and far from convergence

Run for a lot longer... (Which we have done previously but will not do here due to time required)
```{r}

```  

```{r}
## insert your specific directory
load(".rda")
```

```{r}

```  

```{r}
SIRnimble <- nimbleCode({
    ## Initial values
    I[1] <- i0
    S[1] <- pop - i0
    pR <- 1 - exp(-gamma)
    pI[1] <- 1 - exp(-beta*I[1]*S[1]/pop)
    inf[1] ~ dbin(pI[1], S[1])
    infobs[1] ~ dbin(report, inf[1])
    rec[1] ~ dbin(pR, I[1])
    
    ## Step through observations
    for (t in 2:end) {
      S[t] <- S[t - 1] - inf[t - 1]
      I[t] <- I[t - 1] + inf[t - 1] - rec[t - 1]
      pI[t] <- 1 - exp(-beta*S[t]*I[t]/pop)
      inf[t] ~ dbin(pI[t], S[t])
      infobs[t] ~ dbin(report, inf[t])
      rec[t] ~ dbin(pR, I[t])
    }
    ## aux variables
    
    ## priors
    beta ~ dgamma(1/2, 1)
    gamma ~ dunif(0, 1) 
    report ~ dbeta(1/2, 1/2)
  })
```

Set up the model. Here we need:
1) Constants -- 
2) Data -- 
3) Initial Values --
4) NIMBLE model object -- 

```{r}
  SIRcons <- list(i0 = SIRdata[1, 3], 
                     pop = 500, end = 30)
  
  SIRdata <- list(infobs = SIRdata[, 5])
  
  SIRinits <- list(beta = 0.01, gamma = 0.15, report = 0.2,
                     inf = SIRdata[, 5] + 1,
                     rec = round(SIRdata[, 5]/2))
  
  SIRout <- nimbleModel(code = SIRnimble, 
                         name = 'SIR_out', 
                         constants = SIR_Consts,
                         data = SIR_Data, 
                         inits = SIR_Inits)
  

```

```{r}
SIR_out$getNodeNames()
```

```{r}
SIR_out$infobs
```

```{r}
par(mfrow = c(1,1))
tkplot(SIR_out$graph)
```

```{r}
SIR_out$getDependencies(c('beta', 'gamma', 'report'))
```

```{r}
SIR_out$getDependencies(c('beta', 'gamma', 'report'), determOnly = TRUE)
```

```{r}
C_SIR_out <- compileNimble(SIR_out)
```

```{r}
SIR_out_Spec <- configureMCMC(SIR_out, print = TRUE)
```

```{r}
SIR_out_Spec$addMonitors(c('beta', 'gamma', 'report'))
```

```{r}
SIR_out_MCMC <- buildMCMC(SIR_out_Spec)

SIR_out_MCMC <- compileNimble(SIR_out_MCMC, project = SIR_out)
```

```{r}
niter <- 24000
set.seed(0)
SIR_out_MCMC$run(niter)
```

```{r, echo = FALSE}
  samples <- as.matrix(SIR_out_MCMC$mvSamples)
  par(mfrow = c(1, 2), mai = c(.6, .5, .1, .2))
  plot(samples[ , 'beta'], type = 'l', xlab = 'iteration',
       ylab = expression(beta))
  plot(samples[ , 'beta'], type = 'l', xlab = 'iteration',
       ylab = expression(beta), xlim = c(2000, 10000),
       ylim = c(0.00, 0.02))
  plot(samples[ , 'report'], type = 'l', xlab = 'iteration',
       ylab = expression(report))
  plot(samples[ , 'gamma'], type = 'l', xlab = 'iteration',
       ylab = expression(gamma))
```

```{r}
  mc <- MCMCsuite(code = SIRnimble,
                  data = SIR_Data,
                  inits = SIR_Inits,
                  constants = SIR_Consts,
                  MCMCs = c("jags", "nimble", "nimble_slice"),
                  monitors = c("beta", "gamma", "report"),
                  niter = 10000,
                  makePlot = TRUE,
                  savePlot = TRUE)
```

```{r}
  acf(samples[, "beta"])
  acf(samples[, "gamma"])
  acf(samples[, "report"])
```

```{r}
SIR_out_Spec$addSampler(target = c('beta', 'report'), type = 'RW_block',
                      control = list(adaptInterval = 1000))
```

```{r}
SIR_out_Spec$setThin(120)
```

```{r}
SIR_out_MCMC <- buildMCMC(SIR_out_Spec)
```

```{r}
SIR_out_MCMC <- compileNimble(SIR_out_MCMC, project  = SIR_out,
                                resetFunctions = TRUE)
```

```{r}
SIR_out_MCMC$run(niter)
  ## NULL
samplesNew <- as.matrix(SIR_out_MCMC$mvSamples)
```

```{r}
  par(mfrow = c(1, 3), mai = c(.6, .5, .1, .2))
  plot(samplesNew[ , 'beta'], type = 'l', xlab = 'iteration',
       ylab = expression(beta))
  plot(samplesNew[ , 'gamma'], type = 'l', xlab = 'iteration',
       ylab = expression(gamma))
  plot(samplesNew[ , 'report'], type = 'l', xlab = 'iteration',
       ylab = "report")
  par(mfrow = c(1,1))
  plot(samplesNew[ , 'beta'], samplesNew[ , 'gamma'], 
       xlab = expression(beta), ylab = expression(gamma))
```

```{r}
  acf(samplesNew[, "beta"])
  acf(samplesNew[, "gamma"])
  acf(samplesNew[, "report"])
```

```{r}

```