---
title: "Spatial Kalman filter model"
author: "Edgar J. Gonzalez"
date: "November 23, 2015"
output: html_document
---

Install the animation, automap, scatter3d, sp, spacetime, Stem packages.

This package is based on:

Fasso, A. & Cameletti, M. 2010. A unified statistical approach for simulation, modeling, analysis and mapping of environmental data. Simulation: Transactions of the Society for Modeling and Simulation International 86: 139-154.

We have the original Kalman filter formulated as: 

+ *Observation model* 
$$z_{t} = Hy_{t} + \nu_{t},\ \ \ \nu_{t} \sim N(0,\sigma^2_{\nu}I)$$

+ *Process model*
$$y_{t} = Fy_{t-1} + Gx_{t} + \omega_{t},\ \ \ \omega_{t} \sim N(0,\Sigma_{\omega})$$

We modify it by introducing space $s$:

+ *Observation model* 
$$z_{t}(s) = Hu_{t}(s) + \nu_{t}(s),\ \ \ \nu_{t}(s) \sim N(0,\sigma^2_{\nu}I)$$

+ *Process model*
$$u_{t}(s) = K(s)y_{t}(s) + Gx_{t}(s) + \omega_{t}(s),\ \ \ \omega_{t}(s) \sim N(0,\Sigma_{\omega})$$ 
and 
$$y_{t}(s) = Fy_{t-1}(s) + \eta_{t}(s),\ \ \ \eta_{t}(s) \sim N(0,\sigma^2_{\eta})$$

The process starts with a $y_0(s)$ vector taken from a distribution $N(\mu_0,\sigma_0)$.

**Assumptions**

$\omega_{t}(s)$ is separable, so that, $cov(\omega_{t}(s),\omega_{t}(s+h)) = \sigma^{2}_{\omega}\ exp(-\theta h)\ \forall\ t$.

$\nu_{t}(s), \omega_{t}(s)$ and $\eta_{t}(s)$ are independent over time as well as mutually independent.

Therefore, the parameters to be estimated are $\sigma^2_{\nu}$, $\sigma^2_{\omega}$, $\theta$, $G$, $F$,  $\sigma^2_{\eta}$, $\mu_0(s)$ and $\sigma_0$.

**Visualization**

```{r eval = FALSE}
require(animation)
require(automap)
require(scatterplot3d)
require(sp)
require(spacetime)
require(Stem)

data(pm10)
x <- pm10$coords[,1]
y <- pm10$coords[,2]
z <- pm10$z
coordinates <- pm10$coords
covariates <- pm10$covariates
```

Observed process: concentration of particulate matter
```{r eval = FALSE}
saveGIF({
  for(i in 1:nrow(z)) {
    z.i <- z[i,]
    scatterplot3d(x,y,z.i,zlim = c(floor(range(pm10$z)[1]),ceiling(range(pm10$z)[2])), xlab = "UTMX_km", ylab = "UTMX_km", zlab = "log PM10 concentration", type = "h", lty.hplot = 2, pch = 16)
  }
}, movie.name = "PM10.gif", interval = 0.1)
```

Explanatory variable 1 varianble in time: particle primary emissions
```{r eval = FALSE}
saveGIF({
  for(i in 1:nrow(z)) {
    z.i <- covariates[(i-1)*ncol(z)+1:ncol(z), 2]
    scatterplot3d(x,y,z.i,zlim = c(floor(range(covariates[,2])[1]),ceiling(range(covariates[,2])[2])), xlab = "UTMX_km", ylab = "UTMX_km", zlab = "PPM (g/s)", type = "h", lty.hplot = 2, pch = 16)
  }
}, movie.name = "PPM.gif", interval = 0.1)
```
We observe more temporal than spatial variation.

Explanatory variable 2 constant in time: altitude
```{r eval = FALSE}
z.i <- unique(covariates[, 3])
scatterplot3d(x,y,z.i,zlim = range(covariates[, 3]), xlab = "UTMX_km", ylab = "UTMX_km", zlab = "Altitude (km)", type = "h", lty.hplot = 2, pch = 16)
```

**Estimation**

*Null model*

We define the initial values for the parameters  
```{r eval = FALSE}
phi0 <- list(sigma2eps = 0.1, sigma2omega = 0.2, theta = 0.01, beta = matrix(3.65, 1, 1), G = matrix(0.77, 1, 1), Sigmaeta = matrix(0.3, 1, 1), m0 = as.matrix(0), C0 = as.matrix(1))
```

And the $K(s)$ function (identity function here)  
```{r eval = FALSE}
K <- matrix(1, ncol(z), 1)
```

We create the model
```{r eval = FALSE}
mod0 <- Stem.Model(z = z, covariates = as.matrix(covariates[, 1]), coordinates = coordinates, phi = phi0, K = K)
```

Estimate parameters using the EM algorithm
```{r eval = FALSE}
mod0.est <- Stem.Estimation(mod0)
mod0.est$estimates$phi.hat
```

AIC of the null model:
```{r eval = FALSE}
AIC0 <- -2*mod0.est$estimates$loglik + 2*length(mod0.est$estimates$phi.hat)
```

*Full model*  

We define the initial values for the parameters  
```{r eval = FALSE}
phi1 <- list(sigma2eps = 0.1, sigma2omega = 0.2, theta = 0.01, beta = matrix(c(3.65, 0.046, -0.904), 3, 1), G = matrix(0.77, 1, 1), Sigmaeta = matrix(0.3, 1, 1), m0 = as.matrix(0), C0 = as.matrix(1))
```

We create and optimize the model
```{r eval = FALSE}
mod1 <- Stem.Model(z = z, covariates = covariates, coordinates = coordinates, phi = phi1, K = K)
mod1.est <- Stem.Estimation(mod1)
```

And check estimates:  
```{r eval = FALSE}
mod1.est$estimates$phi.hat
```

AIC of the full model:
```{r eval = FALSE}
AIC1 <- -2*mod1.est$estimates$loglik + 2*length(mod1.est$estimates$phi.hat)
```

**Prediction (Kriging)**  

*Null model*
```{r eval = FALSE}
n.times <- 10
n.grid <- 25
Krig.x <- seq(400, 470, length = n.grid)
Krig.y <- seq(5000, 5070, length = n.grid)
Krig.n <- n.grid*n.grid
Krig.coord <- expand.grid(x = Krig.x, y = Krig.y)
Krig.covar <- cbind(rep(1,Krig.n))
Krig.K <- matrix(1, Krig.n, 1)
saveGIF({
  for(i in 1:n.times) {
    mod0.pred <-Stem.Kriging(StemModel = mod0.est, coord.newlocations = Krig.coord, covariates.newlocations = Krig.covar, K.newlocations = Krig.K, time.point = i)
    z.i <- matrix(mod0.pred$data.newlocations$z, n.grid, n.grid)
    persp(Krig.x, Krig.y, z.i, zlab = "", zlim = range(z), axes = FALSE)
  }
}, movie.name = "PM10_kriging0.gif", interval = 0.3)
```

*Full model*

We want 10 time units over a grid:
```{r eval = FALSE}
n.times <- 10
n.grid <- 25
Krig.x <- seq(340, 490, length = n.grid)
Krig.y <- seq(4900, 5090, length = n.grid)
Krig.n <- n.grid*n.grid
Krig.coord <- expand.grid(x = Krig.x, y = Krig.y)
```

First we have to interpolate the explanatory variables that remain constant in time (altitude):
```{r eval = FALSE}
alt.x <- coordinates[,1]
alt.y <- coordinates[,2]
alt.s <- as.data.frame(cbind(alt.x,alt.y))
coordinates(alt.s) =~ alt.x+alt.y
alt.s$alt <- unique(covariates[,3])
coordinates(Krig.coord) =~ x+y
alt.mod <- autoKrige(alt.s$alt~1, alt.s, Krig.coord) 
alt.new <- alt.mod$krige_output$var1.pred
res <- persp(Krig.x, Krig.y, matrix(alt.new, n.grid, n.grid), zlab = "", axes = FALSE)
z.i <- unique(covariates[, 3])
points <- trans3d(x,y,z.i, pmat = res)
points(points, pch = 16)
```

Concerning the explanatory variable 1 (particle primary emissions) since most of the variation is in time and in that axis variable 1 only changes the slope of the graph, we'll plot a low PPM value (50 g/s)
```{r eval = FALSE}
Krig.coord <- expand.grid(x = Krig.x, y = Krig.y)
Krig.covar <- cbind(rep(1,Krig.n), rep(50,Krig.n), alt.new)
Krig.K <- matrix(1, Krig.n, 1)
saveGIF({
  for(i in 1:n.times) {
    mod1.pred <- Stem.Kriging(StemModel = mod1.est, coord.newlocations = Krig.coord, covariates.newlocations = Krig.covar, K.newlocations = Krig.K, time.point = i)
    z.i <- matrix(mod1.pred$data.newlocations$z, n.grid, n.grid)
    persp(Krig.x, Krig.y, z.i, zlab = "", zlim = range(z), axes = FALSE)
  }
}, movie.name = "PM10_kriging1_50.gif", interval = 0.3)
```