---
title: "Spectral model"
author: "Edgar J. Gonzalez"
date: "November 23, 2015"
output: html_document
---

Install the spate and animation package

We can specify a Gaussian process through a stochastic partial differential equation (SPDE), which implicitly gives a covariance function. The advection-diffusion SPDE is a basic model. 

$$\frac{\partial }{\partial t}\xi(t,s)= -\mu\frac{\partial }{\partial s}\xi(t,s) + \triangledown\cdot\Sigma\frac{\partial }{\partial s}\xi(t,s) - \zeta\xi(t,s) + \epsilon(t,s)$$

where

$\triangledown\cdot F$ is the divergence operator (which describes the outward sink of the field).

$$\Sigma^{-1} = \frac{1}{\rho^2_1}
\left( \begin{array}{ccc}
\cos\alpha & \sin\alpha \\
-\gamma\cdot\sin\alpha  & \gamma\cdot\cos\alpha\end{array} \right)^T\left( \begin{array}{ccc}
\cos\alpha & \sin\alpha \\
-\gamma\cdot\sin\alpha  & \gamma\cdot\cos\alpha\end{array} \right)$$

and $\epsilon(t,s)$ is a Gaussian random field.

An SPDE can be discretized in space and time and solved in the spectral space.

We thus use a Fourier-transformed spatio-temporal field:

$$\sum_{k = 1}^{m}\phi_{k}^{(c)}(s)\beta_{j}^{(c)}(t) + \sum_{k = 1}^{m}\phi_{k}^{(s)}(s)\beta_{j}^{(s)}(t)$$

where $\phi_{k}^{(c)}(s) = \cos(k'_{j}s)$ and $\phi_{k}^{(s)}(s) = \sin(k'_{j}s)$ are a Fourier basis.

And have the model:

+ *Observation model* 
$$z_{t+1}(s) = y_{t+1}(s) + \nu_{t+1}(s),\ \ \ \nu_{t}(s) \sim N(0,\sigma^2_{\nu}I) \ \forall t$$

+ *Transformation to the Fourier space*
$$y_{t+1}(s) = \Phi\alpha_{t+1}(s)$$

+ *Process model*
$$\alpha_{t+1}(s) = G\alpha_{t}(s) + q_{t+1},\ \ \ 
q_{t} \sim N(0,Q)$$

where $G$ is called the propagator function (matrix) and $Q$ is called the innovation covariance matrix, which are basically the equivalents of the $K(s)$ and $\Sigma_w$ in the non-transformed models.

We now use the package:

**Simulating data**

We simulate the observed process over 20 places and 4 time units:
```{r eval = FALSE}
require(spate)
n <- 20
Time <- 4
par <- c(rho0 = 0.1, sigma2 = 0.2, zeta = 0.5, rho1 = 0.1, gamma = 2,
         alpha = pi/4, muX = 0.2, muY = -0.2, tau2 = 0.01)
par0 <- c(rho0 = 0.1, sigma2 = 0.2, zeta = 0.5, rho1 = 0.1, gamma = 2,
         alpha = pi/4, muX = 0.2, muY = -0.2, tau2 = 0.01)
spateSim <- spate.sim(par = par, n = n, T = Time, seed = 4)
```

**Visualization**
```{r eval = FALSE}
plot(spateSim, mfrow = c(1,4), mar = c(2,2,2,2), indScale = TRUE, cex.axis = 1.5, cex.main = 2)
```

**Estimation**

We simulate over a little more time units...
```{r eval = FALSE}
n <- 20
Time <- 20
spateSim <- spate.sim(par = par, n = n, T = Time, seed = 1000)
```

Extract the observed process:
```{r eval = FALSE}
w <- spateSim$w
```

And Fourier-transform them:
```{r eval = FALSE}
wFT <- real.fft.TS(w,n = n,T = T)
```

Lets not start at the same parameter values:
```{r eval = FALSE}
parI <- c(rho0 = 0.2, sigma2 = 0.1, zeta = 0.25, rho1 = 0.01, gamma = 1, alpha = 0.3, muX = 0, muY = 0, tau2 = 0.005)
```

A detail: weird way to say which parameters are in log scale.
```{r eval = FALSE}
logInd=c(1,2,3,4,5,9)
parI[logInd] <- log(parI[logInd])
```

And we estimate using standard optim():
```{r eval = FALSE}
spateMLE <- optim(par = parI, loglike, control = list(trace = TRUE, maxit = 1000), wFT = wFT, method="L-BFGS-B", lower = c(-10,-10,-10,-10,-10,0,-0.5,-0.5,-10), upper = c(10,10,10,10,10,pi/2,0.5,0.5,10), negative = TRUE, logScale = TRUE,logInd = c(1,2,3,4,5,9), hessian = TRUE,n = n,T = T)
spateMLE$par
```

**MCMC**
If we want to sample the a posteriori parameter distributions, we can use the authors' function spate.mcmc():
```{r eval = FALSE}
spateMCMC <-spate.mcmc(y = w, x = NULL, SV = c(rho0 = 0.2, sigma2 = 0.1, zeta = 0.25, rho1 = 0.2, gamma = 1, alpha = 0.3, muX = 0, muY = 0, tau2 = 0.005), RWCov = diag(c(0.005,0.005,0.05,0.005,0.005,0.001,0.0002,0.0002,0.0002)), Nmc = 10000, BurnIn = 2000, seed = 4, NCovEst = 500, BurnInCovEst = 500, trace = FALSE)
spateMCMC
```

**Prediction**

```{r eval = FALSE}
#If you skip the MCMC you can use the spateMCMC data
data("spateMCMC")
predict <- spate.predict(y = w, tPred = c(17:50), spateMCMC = spateMCMC, Nsim = 200, BurnIn = 10)
Pmean <- apply(predict, c(1,2), mean)
Psd <- apply(predict,c(1,2),sd)
```

Now we compare observed and fitted fields
```{r eval = FALSE}
par(mfrow = c(2,2))
image(1:20, 1:20, matrix(w[19, ], nrow = 20), main = "Obs. field at t = 19", xlab = "x", ylab = "y")
image(1:20, 1:20, matrix(Pmean[3, ], nrow = 20), main = "Fit. field at t = 19", xlab = "x", ylab = "y")
image(1:20, 1:20, matrix(w[20, ], nrow = 20), main="Obs. field at t = 20", xlab = "x", ylab = "y")
image(1:20, 1:20, matrix(Pmean[4, ], nrow = 20), main = "Fit. field at t = 20", xlab = "x", ylab = "y")
```

And an animation of predictions...
```{r eval = FALSE}
zlim = c(min(Pmean), max(Pmean))
saveGIF({
  for(i in 1:34) {
image(1:20,1:20,matrix(Pmean[i,],nrow=20), zlim = zlim,
main = paste0("t = ",i+16), xlab = "x", ylab = "y")
  }
}, movie.name = "spectral.gif", interval = 0.4)
```
