---
title: "part3"
author: "Alex Kuzmin and Joan Lopez"
date: '2018-03-26'
output: html_document
---

### Part 3 - Other graphics 

### Example 3.1 - Chord diagrams 

This is used to show inter-relationships between data in matrix. The data are arranged around a circle and the relationships between the data are the arcs that go from one end of the circle to another. 

Here's a small example: 


```{r}
library(chorddiag)

students = data.frame(Math = c(50, 25, 5, 12),
                      Art = c(10, 55, 5, 20),
                      Science = c(45,12,29, 20),
                      PE = c(24,67,27,15))

students = as.matrix(students)
row.names(students) = c("Section A", "Section B", "Section C", "Section D")

chorddiag(students, 
          type = "bipartite", 
          palette = "YlOrRd",
          showTicks = F, 
          groupnameFontsize = 14, 
          groupnamePadding = 10, 
          margin = 90
          )
```

"type" refers to the type of chord diagram. can be either "directional" or "bipartite"
"palette" is the name of the colourbrewer palette. 

<br>
<br>


### Example 3.2 - interactive heatmap 

```{r}
library(d3heatmap)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv"
drivers <- read.csv(url, row.names = 1)
d3heatmap(drivers, 
          scale = "column",
          colors = "YlGnBu",
          #dendrogram = "none",
          #Rowv = FALSE,
          #Colv = FALSE
          dendrogram = "row", 
          k_row = 3,
          xaxis_font_size = "6pt"
          )
```

<br>
<br>
<br>

This dataset was obtained from FiveThirtyEight and they were looking at different variables to see if there was a way to show which state had the worst drivers.

<br>
<br>
<br>

For the assignment: replicate one of the graphs/make your own plot OR find an example online and comment on the lines of code (R code).  


