---
title: "Parallel Computing"
author: "Jennifer La Rosa, Laxman Ghimire"
date: "November 16, 2015"
output: html_document
---

###Parallel Computing

Due to the chains being independent, we can use parallel computing to reduce the computation time of data cloning problems. This code comes from (1). 

```{r setup}
library(dclone)
library(rjags)
library(coda)
library(dcmle)
if (isTRUE(getOption('knitr.in.progress'))) {
    ## turn off progress bars
    dcoptions(verbose=0)
    options(jags.pb="none")
}
##load("solymos.rda")  ## poor man's caching

## my specific directory, see above for general
load("C:/Users/Owner/Downloads/solymos (2).rda")
```

```{r}
set.seed(4321) # set random seed for reproducibility
n <- 25 # sample size
p <- 0.3 # true parameter value
y <- rbinom(n = n, size = 1, prob = p)
```

```{r}
## dclone-ified model specification
modelpc <- custommodel("model {
                       for (k in 1:K) {
                       for (i in 1:n) {
                       Y[i,k] ~ dbin(p, 1)
                       }
                       }
                       p ~ dunif(0.001, 0.999)
                       }")

## dclone-ified data specification
datpc <- list(Y = dcdim(data.matrix(y)), n = n, K = 1)

## parallel data cloning based MCMC results
cl <- makeCluster(3)
dcpfit <- dc.parfit(cl, data = datpc, params = "p", model = modelpc,
                    n.clones = c(1,2,4,8), unchanged = "n", multiply = "K",
                    n.chains = 3, partype = "parchains")
stopCluster(cl)

summary(dcpfit)

coef(dcpfit) # MLE
dcsd(dcpfit) # asymptotic SEs
vcov(dcpfit) # inverse Fisher information matrix
confint(dcpfit) # asymptotic confidence interval