


---
title: "Kalman Filter lab"
author: "Sina Brar and Athinthtra Krishnaswamy"
date:  "`r format(Sys.time(), '%H:%M %d %B %Y')`"
bibliography: nonlin.bib
---

## Nile's annual flow

```{r message=FALSE}
library("FKF")
library("dlm")
```

The `Nile` data set is built into R (it's in the `datasets` recommended
package) - it doesn't need to be loaded explicitly.


## Lab Activity

Read through the following code, to see an example of Kalman filtering
using the `dlm` package.

Afterwards, try doing this example using the `FKF` package.
Read through the `FKF` documentation either [online](https://cran.r-project.org/web/packages/FKF/FKF.pdf) or via `help(package="FKF")`.

Then, use either the `dlm` or `FKF` package to do Kalman filtering on the 
[nhtemp]() or [JohnsonJohnson]() datasets in R, using the same model.

Commands to install packages:
```{r install_pkgs,eval=FALSE}
install.packages("FKF")
install.packages("dlm")
```

## dlm: Dynamic Linear Model

### dlm package example using Nile Flow data

An object of class `dlm` is created through the dlm function:
```{r dlmex,eval=FALSE}
dlm(m0, C0, FF, V, GG, W)
```
where

* `V`: covariance matrix of observation equation
* `FF`: transition matrix of observation equation
* `W`: covariance matrix of state equation 
* `GG`: transition matrix of state equation (i.e., prediction 
* `m0`, `C0`: initial values of mean and variance-covariance matrix

The `dlmModReg`, `dlmModPoly`, `dlmModARMA`, and `dlmModSeas`
functions   create particular objects of class `"dlm"`;
see the documentation for further details

For this example, we have the following model:

```
x[t]  = GG * x[t-1] + w[t]
z[t]  = FF * x(t) + v(t)
```


with `FF == GG == 1`. `W` is the covariance of `w[t]` the process noise;
`V` is the covariance of `v[t]`, the observation noise.

* **Q**: what kind of dynamics are we assuming with this model???

1. We need to find values for `GG`, `FF`, `V`, `W`, `m0`, `C0` in order to apply KF
2. `FF` and `GG` are already known, `m0` and `C0` receive default values
3. We'll use maximum likelihood estimation to find values for `W` and `V`


### Finding the MLE of W and V


Note: `dlm` has a specific function in the package for this.
Since FKF only has two functions, one can use optim to do 
essentially the same thing. The function `optim` optimizes is
the loglikelihood which is one of the values returned by the 
`fkf` function. See the FKF documentation for more details or
command `?fkf` in the shell

`dlm` uses the function `dlmMLE`: 

```{r eval=FALSE}
 dlmMLE(y, parm, build,...)
```
where

* `y`: data
* `parm`: initial values for parameters that need to be estimated
* `build`: function that takes a vector of the same length as `parm`, and returns object of class `dlm`
It returns the MLE of unknown parameters in the specification of a state space model.
The unknown parameters in this case are `V` and `W`

`dlmMLE` takes a function that returns a `dlm` object as one of its arguments.
It then gives us the estimates of the unknown parameters in the model 
specified by that `dlm` object. 
So we need a function that will create our `dlm` object/model:

`nileBuild` creates a dlm with FF = 1, GG = 1, m0 = 0, C0=1e+07, 
and variance matrices as specified by `par`, a vector of length 2.
      
```{r nileBuild}	  
nileBuild <- function(par) {
    dlmModPoly(1, dV = exp(par[1]), dW = exp(par[2]))
}
```
      
`dlmModPoly` creates an nth order `dlm` model. The default values it takes 
are as follows:

```{r dlmModPoly,eval=FALSE}
dlmModPoly(order=2, dV = 1, dW = c(rep(0, order - 1), 1),
       m0 = rep(0, order), C0 = 1e+07 * diag(nrow = order))
```

It returns an object of class `dlm`
You could also use the `dlm()` function and specify the known values 
for the arguments.

Call `dlmMLE`:

```{r dlmMLE}
nileMLE <- dlmMLE(Nile, rep(0,2), nileBuild)
nileMLE$par #vector containing estimates of V and W
```

Now use the MLEs found in above step to create a `dlm` object (model)

```{r nileMod}
(nileMod <- nileBuild(nileMLE$par))
```

Apply the Kalman Filter to our model: `dlmFilter(y, mod)` with
data `y` and a `mod` object of class `dlm`
applies Kalman filter to compute filtered values of the state
vectors, together with their variance/covariance matrices.

```{r runKF}
nileFilt <- dlmFilter(Nile, nileMod)
```
call `?dlmFilter` to see what it returns 

Plot the flow data together with filtered values of the state vectors:

```{r plot}
plot(Nile, main = "Nile flow")
tvec <- c(time(Nile))  ## extract time vector, drop time-series attributes
lines(x=tvec,y=nileFilt$m[-1], col = "red")
```


